---
title: "advent of code - day 2"
author: "chad allison | 2 december 2022"
output: github_document
---

### loading in tidyverse

```{r message = F, warning = F}
library(tidyverse)
```

### loading in input file

```{r}
input = readLines("day2_input.txt")
```

### previewing input file

```{r}
input[1:100]
```

### formatting the input into a data frame to manipulate

```{r}
df = data.frame(input) |>
  separate(input, into = c("opponent", "me"), sep = " ")

head(df)
```

### converting letters to choices for quality of life

```{r}
df = df |>
  mutate(opponent_choice = case_when(opponent == "A" ~ "rock",
                                     opponent == "B" ~ "paper",
                                     opponent == "C" ~ "scissors"),
         me_choice = case_when(me == "X" ~ "rock",
                               me == "Y" ~ "paper",
                               me == "Z" ~ "scissors"))

head(df)
```

### creating winner variable

```{r}
df = df |>
  mutate(winner = case_when(opponent_choice == "paper" & me_choice == "paper" ~ "draw",
                            opponent_choice == "paper" & me_choice == "rock" ~ "opponent",
                            opponent_choice == "paper" & me_choice == "scissors" ~ "me",
                            opponent_choice == "rock" & me_choice == "paper" ~ "me",
                            opponent_choice == "rock" & me_choice == "rock" ~ "draw",
                            opponent_choice == "rock" & me_choice == "scissors" ~ "opponent",
                            opponent_choice == "scissors" & me_choice == "paper" ~ "opponent",
                            opponent_choice == "scissors" & me_choice == "rock" ~ "me",
                            opponent_choice == "scissors" & me_choice == "scissors" ~ "draw"))

head(df)
```

### calculating score

```{r}
df = df |>
  mutate(score = 0,
         score = case_when(me_choice == "rock" ~ score + 1,
                           me_choice == "paper" ~ score + 2,
                           me_choice == "scissors" ~ score + 3),
         score = case_when(winner == "draw" ~ score + 3,
                           winner == "me" ~ score + 6,
                           T ~ score))

head(df)
```

### part 1 solution

```{r}
sum(df$score)
```

### reverting to older version of data

```{r}
df = data.frame(input) |>
  separate(input, into = c("opponent", "outcome"), sep = " ") |>
  mutate(opponent_choice = case_when(opponent == "A" ~ "rock",
                                     opponent == "B" ~ "paper",
                                     opponent == "C" ~ "scissors"))

head(df)
```

### creating desired outcome variable

```{r}
df = df |>
  mutate(desired_outcome = case_when(outcome == "X" ~ "lose",
                                     outcome == "Y" ~ "draw",
                                     outcome == "Z" ~ "win"))

head(df)
```

### deciding what shape we need to choose for desired outcome

```{r}
df = df |>
  mutate(me_choice = case_when(desired_outcome == "draw" ~ opponent_choice,
                               desired_outcome == "lose" & opponent_choice == "rock" ~ "scissors",
                               desired_outcome == "lose" & opponent_choice == "paper" ~ "rock",
                               desired_outcome == "lose" & opponent_choice == "scissors" ~ "paper",
                               desired_outcome == "win" & opponent_choice == "rock" ~ "paper",
                               desired_outcome == "win" & opponent_choice == "paper" ~ "scissors",
                               desired_outcome == "win" & opponent_choice == "scissors" ~ "rock"))

head(df)
```

### recreating winner variable and calculating new total score

```{r}
df = df |>
  mutate(winner = case_when(opponent_choice == "paper" & me_choice == "paper" ~ "draw",
                            opponent_choice == "paper" & me_choice == "rock" ~ "opponent",
                            opponent_choice == "paper" & me_choice == "scissors" ~ "me",
                            opponent_choice == "rock" & me_choice == "paper" ~ "me",
                            opponent_choice == "rock" & me_choice == "rock" ~ "draw",
                            opponent_choice == "rock" & me_choice == "scissors" ~ "opponent",
                            opponent_choice == "scissors" & me_choice == "paper" ~ "opponent",
                            opponent_choice == "scissors" & me_choice == "rock" ~ "me",
                            opponent_choice == "scissors" & me_choice == "scissors" ~ "draw"),
         score = 0,
         score = case_when(me_choice == "rock" ~ score + 1,
                           me_choice == "paper" ~ score + 2,
                           me_choice == "scissors" ~ score + 3),
         score = case_when(winner == "draw" ~ score + 3,
                           winner == "me" ~ score + 6,
                           T ~ score))

head(df)
```

### part 2 solution

```{r}
sum(df$score)
```
















