---
title: "advent of code - day 6"
author: "chad allison | 6 december 2022"
output: github_document
---

### loading tidyverse

```{r message = F, warning = F}
library(tidyverse)
```

### loading input data

```{r}
input = readLines("day6_input.txt")

input
```

### creating empty data frame to track each marker

```{r}
df = data.frame(marker = rep(NA, times = nchar(input) - 3), marker_num = NA, n_processed = NA)

head(df)
```

### creating function to check if four characters are all different

```{r}
check_diff = function(x) {
  
  return(substr(x, 1, 1) != substr(x, 2, 2) & substr(x, 1, 1) != substr(x, 3, 3) &
           substr(x, 1, 1) != substr(x, 4, 4) & substr(x, 2, 2) != substr(x, 3, 3) &
           substr(x, 2, 2) != substr(x, 4, 4) & substr(x, 3, 3) != substr(x, 4, 4))
  
}

check_diff("abcc")
check_diff("abcd")
```

### filling data frame with each four-character marker

```{r}
for (i in 1:nrow(df)) {
  
  df$marker[i] = substr(input, i, i + 3)
  df$marker_num[i] = i
  df$n_processed[i] = i + 3
  
}

head(df)
```

### applying function to data

```{r}
df = df |>
  mutate(all_different = check_diff(marker))

head(df)
```

### part 1 solution

```{r}
df |>
  filter(all_different) |>
  pull(n_processed) |>
  head(1)
```

### recreating data frame with all possible markers

```{r}
df = data.frame(marker = rep(NA, times = nchar(input) - 13), marker_num = NA, n_processed = NA)

head(df)
```

### creating function to check if all fourteen characters are different

```{r}
check_diff2 = function(x) {
  
  count = data.frame(letters = str_split(x, pattern = "")[[1]]) |>
    count(letters) |>
    nrow()

  return(count == nchar(x))
  
}

check_diff2("abcdefghijklmm")
check_diff2("abcdefghijklmn")
```

### filling data frame with each fourteen-character marker

```{r}
for (i in 1:nrow(df)) {
  
  df$marker[i] = substr(input, i, i + 13)
  df$marker_num[i] = i
  df$n_processed[i] = i + 13
  
}

head(df)
```

### applying function to data

```{r}
df = df |>
  rowwise() |>
  mutate(all_different = check_diff2(marker))

head(df)
```

### part 2 solution

```{r}
df |>
  filter(all_different) |>
  pull(n_processed) |>
  head(1)
```





















